Replacing and additive horizontal \\ gene transfer in {\em Streptococcus}

### 2011-12-08 doc/manuscript.tex
Figures are changed. The manuscript is in the tex file.

% Four potential reviewers whose e-mail addresses you can provide
% Debra Bessen
% Claus O. Wilke
% Didelot Xavier
% Bruce R. Levin
% Michael Wessels

% Template for PLoS
% Version 1.0 January 2009
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template

\documentclass[10pt]{article}

@@

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

@@

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

@@

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

@@

% Use the PLoS provided bibtex style
% \bibliographystyle{plos}

@@

% FIXME: Remove the followings if you have to when submitting the manuscript.
\let\citep\cite
\let\citet\cite
\providecommand{\tabularnewline}{\\}
\newcommand{\lyxdot}{.}
\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}

@@

\pagestyle{myheadings}
%% ** EDIT HERE **
\markright{Gene Transfer using Streptococcus Genomes}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
%% END MACROS SECTION

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Inferring homologous recombination and horizontal gene transfer 
using whole genomes of \textit{Streptococcus pyogenes} 
and\textit{ S.\ dysgalactiae} ssp.\textit{\ equisimilis}}
}
% Short Title (running head): Gene transfer using Streptococcus genomes
% Insert Author names, affiliations and corresponding author email.
\\
Sang Chul Choi$^{1}$, Matthew D. Rasmussen$^{1}$, 
Melissa Jane Hubisz$^{1}$, Ilan Gronau$^{1}$,
Michael J. Stanhope$^{2}$, and Adam Siepel$^{1,\ast}$
\\
\bf{1} 
Department of Biological Statistics and Computational Biology,
Cornell University, Ithaca, NY 14853, USA
\\
\bf{2}
Department of Population Medicine and Diagnostic Sciences,
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\
$\ast$ E-mail: acs4@cornell.edu, Phone: 607-254-1157, Fax: 607-255-4698
\end{flushleft}

% \section*{Things To Do for submitting this to PLoS Pathogens}
% \begin{enumerate}
% \item Legends for the main figures as a separate section
% \item We can add at least 50 words in section \textbf{Abstract}.
% \item Figures, tables, and supporting information need to be ready.
% \item Merge Genome Browser tracks to the official one.
% \item Publish source codes at public domain.
% \end{enumerate}
%
% \section*{A list of our major and minor findings or observations}
% \begin{enumerate}
% \item We showed that recombining gene transfer from SPY to SDE was strong
% enough.
% \item A few evidences for the net directionality from SPY to SDE:
% time-constrained gene transfer between SPY to SDE also showed the directionality
% from SPY to SDE.
% \item We showed that virulence genes were associated with horizontal gene
% transfer. 
% \item We found that recombining gene tranfer processes are involved in
% membrane-bound protein encoding genes and aminoacyl tRNA synthetases.
% \item We found that horizontal gene transfer processes are involved in
% transposases.
% \item We provided evidence that recombination process affected more sites along
% the core genome of SPY and SDE than mutation process.
% \item We found a recombination hotspot at 200 kb from the origin of replication
% along the genome of SDE1.
% \item We showed that species-tree-topology-changing recombinant edges were more
% reliably recovered than species-tree-topology-not-changing edges.
% \item We showed that recombination intensity might be useful in finding genes
% under high recombining gene transfer.
% \item We found that SPY and SDE are more closely related than SDD and SDE in
% about 10\% of the core genome.
% \item Recombination intensity and number of recombining gene transfer events
% were positively correlated.
% \item We found reduction of numbers of genes after divergence of SDE and SPY.
% \item We confirmed the species tree that relates the five genomes.
% \item We listed several virulence genes associated with recombining and
% horizontal gene transfer.
% \end{enumerate}
% 
% \section*{Suggestions about the manuscript}
% \begin{enumerate}
% \item Ilan thought that a species tree must be shown at the outset of the
% section \textbf{Results}.
% \item Can selections after gene transfer be tested?
% \end{enumerate}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Two mechanisms of gene transfer in bacteria
were known; recipient bacteria of transferred genes can substitute transferred
genes for homologous parts of their genomes (Recombining Gene Transfer), or
integrate transferred genes into their genomes without replacement (Horizontal
Gene Transfer).  We discriminated the two processes by using a model-based
approach for studying recombining gene transfer, and using a parsimony-based
approach for both transfer processes.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more often
for the net direction from SPY to SDE.  We also found evidence that virulence
genes transferred more often via non-recombining fashion without replacing parts
of destination genomes than did non-virulence genes.  
A genome-wide resource of
recombining and non-recombining gene transfer for the study of the SPY and SDE
genomes is provided so that research community in the pathogenic SPY and SDE
bacteria could take advantage of it.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\textit{Streptococcus pyogenes} (SPY) has afflicted people around the globe
causing many diseases. Its closely related species, \textit{S.\ dysgalactiae}
subspecies \textit{equisimilis} (SDE), used to be known as veterinary pathogens
have been reported to share disease profiles with SPY.  While gene transfers
between the two species were shown in small portions of genomic loci, no study
on the gene transfer at genome-scale has been performed because of lack of
methodologies implemented or genomic data available. Recent advances in both
method development and genomic data allowed to study gene transfer between the
two species in genome scale.  Recipient bacteria of transferred genes can
substitute transferred genes for homologous parts of their genomes, or integrate
transferred genes into their genomes without replacement.  The former is called
recombining gene transfer, and the latter horizontal gene transfer. By
discriminating the two processes, We found that virulence genes were more
associated with horizontal gene transfer than recombining gene transfer.  We
also found a signficant signal of recombining gene transfer for the net
direction from SPY to SDE in the core genome.  Our study provides new insights
into gene transfer between SPY snd SDE in genome scale.

@@

Bacteria infect humans with a variety of diseases: e.g., food poisoning
(\textit{Eschericha Coli}), strep throat (\textit{Streptococcus pyogenes}),
tuberculosis (\textit{Mycobacterium tuberculosis}), cholera (\textit{Vibrio
cholera}), anthrax (\textit{Bacillus anthracis}), pneumonia
(\textit{Chlamydophila pneumoniae}), and influenza (\textit{Haemophilus
influenzae}), to name a few.  \textit{Streptococcus pyogenes} (SPY) belonging to
group A streptococci (GAS) primarily infect humans in the throat and skin
\citep{Broyles2009}. It has high prevalence throughout the world causing
diseases ranging from mild illnesses such as pharyngitis to severe invasive
streptococcal diseases including necrotizing fasciitis and toxic shock syndrome
\citep{Cunningham2000a}.  Its close relative, \textit{Streptococcus
dysgalactiae} subspecies \textit{equisimilis} (SDE) belonging to group C or G
streptococci (GCS/GGS), used to be generally regarded as a veterinary pathogen
\citep{Vandamme1996}. However, it has been reported to cause serious and
life-threatening streptococcal diseases to humans traditionally associated with
GAS \citep{Brandt2009}.  Because of the overlap in disease profile and
ecological niche of tissue sites betwen SPY and SDE, these bacteria may as
well interact with each other by exchanging parts of their genomes.

Therefore, the research topic on gene exchange between SPY and SDE has been
performed in various ways. Sachse et al.\ \citet{Sachse2002} showed gene transfer from SDE to
SPY for the downstream of the genomic region of streptococcal pyogenic
exotoxins. Davies et al.\ \citet{Davies2007} investigated phage 3396 from SDE as a vehicle of
gene transfer from SDE to SPY. They \citet{Davies2009} also demonstrated that an
integrative conjugative element of SDE allowed gene transfer from SDE to other
streptococci. Ahmad et al.\ \citet{Ahmad2009} did not conclusively deduce the net
directionality of housekeeping gene transfer between SPY and SDE.  Notably,
Kalia et al.\ \citet{Kalia2001} attempted to show that the net direction of gene transfer from
SPY to SDE was more dominant than the other direction.  Because these researches
on the controversial issue, net directionality of gene transfer between SPY and
SDE, have been limited to relatively small regions in the genomes primarily on
representative loci from the Multi Locus Sequence Typing (MLST) database, it
remains to be elusive to determine the genome-wide degrees to which genomic
segments flowed between the two species, and net directions of thereof.

As sequencing costs plummet \citep{Mardis2011}, bacterial genome sequencing
produces individual genomes of high quality even within a bacterial species
\citep{Tettelin2009a}.  Studies on horizontal gene transfer in prokaryotes have
used genes in very large sets of species from a great deal of genomes available 
\citep{Koonin2001}.  Although this type of study gave us insights into deep
evolutionary histories of prokaryotes, with multiple individual genomes per
species we now can go beyond such rough broad study on gene transfer towards
fine-scaled history of gene transfer between very closely related prokaryotes.
For instance, Liu et al.\ \citet{Liu2009} studied gene transfer events in
\textit{Lactobacillus bulgaricus} and \textit{Streptococcus thermophilus} using
the compositional approach of Karlin \citet{Karlin2001}. Luo et al.\ \citet{Luo2011} investigated
gene transfer events in \textit{Eschericha Coli} using the embedded quartet
decomposition due to Zhaxybayeva et al.\ \citet{Zhaxybayeva2006}. Hamady et al.\ \citet{Hamady2006} combined
phylogenetic and compositional approaches to detecting horizontal gene transfer.
Caro-Quintero et al.\ \citet{Caro-Quintero2011} studied a number of \textit{Shewanella baltica}
genomes to identify the genetic elements that enabled the species to adapt to
redox gradients.  Although several complete genomes from SPY have been
available, Suzuki et al.\ \citet{Suzuki2011} just recently added one more complete genome of
SDE to the set of a previously unique SDE genome \citep{Shimomura2011}. These
multiple SDE genomes prompted us to study genome-wide gene flows between SPY and
SDE that afflicted human beings around the globe.

Bacteria evolves new biological features through horizontally transferred genes
from other lineages and vertically transmitted mutations on genes within the
same lineage.  While mutations on a gene modifiy relatively small portions of the 
gene through substitutions, insertions, and deletions of deoxyribonucleic acids,
gene transfers can reshape the organization of a bacterial genome,
which could revamp biological functions of the species.  Recipient bacteria of
transferred genes can substitute transferred genes for homologous parts of their
genomes, or integrate transferred genes into their genomes without replacement.
These two processes are here in this manuscript discriminated by calling the
former \textit{recombining gene transfer}, and the latter \textit{horizontal
gene transfer} \citep{Ochman2001,Lawrence2009}.  Despite the two distinct
evolutionary processes in bacteria, analyses of complete genome sequences in
efforts of better understanding bacterial evolution have paid less attention to
a combined approach to illuminating disparity of the two gene transfer
processes. We attempted here to use and compare two approaches for
distinguishing the recombining gene transfer process from the horizontal, or
non-recombining, gene transfer, and tackled the two processes in association with
virulence genes.

Recently, a method of inferring homologous recombination with whole bacterial
genomes was developed \citep{Didelot2010}. The model of homologous recombination
allowed them to study recombining gene flows between bacteria individuals of
\textit{Bacillus cereus} species in the core genomes.   In the model, a species
tree that relates individuals genomes is augmented by attaching
\textit{recombinant edges} between species tree branches that can be used to
measure gene transfers.  Other evolutionary processes acting on bacterial
genomes include genome rearragement, translocation, inversion, duplication,
gain, loss, etc.  These processes did not readily lend themselves to
statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by Doyon et al.\ \citet{Doyon2011}.  In the
approach, gene trees are inferred for gene families that are constructed from
all of the genomes, and are reconciled with a speices tree to infer
parsimoniously events.  We noted that the first approach could be used to infer
transfer of genomic segments due to homologous recombination, and the second
could be useful in studying horizontal gene transfer.

In short, the complete genomes of the two species of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer.  We found that virulence genes were more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parsimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of the core genomes also showed for
the first time that the net directionality of recombining gene transfer from SPY
to SDE was stronger at the genome scale.

% Results and Discussion can be combined.
\section*{Results}

% \texttt{We show a net directionality of recombining gene transfers from SPY to SDE. 
% Describe our arguments of the net directionality. 
% Number of recombinant edges.
% Confidence about the numbers of recombinant edges.
% Not all of genes experience recombining gene transfers from SPY to SDE.
% Gene transfer counts from the parsimony-based approach; the power of the
% parsimony-based approach may be limited.
% Is this number of recombinant edges good enough for the argument?
% Reflection of the previous analyses, 
% heat maps of real data analysis, 
% and supporting simulations.}

\subsection*{Recombining gene transfer from SPY to SDE at genome-wide scale}

We followed the pipeline and methods recommended by Didelot et al.
\cite{Didelot2010} to pursue a model-based study of recombining gene transfer. A
preliminary step in the pipeline involved estimating the species tree with
branch lengths via \texttt{ClonalFrame} \cite{Didelot2007}, and three population
parameters of mutation rate, recombination rate, and average recombinant tract
length via a first phase of \texttt{ClonalOrigin} \cite{Didelot2010}.  Because
\textit{S.\ equi} ssp.\textit{\ equi} strain 4047 (SEE) diverged significantly
from the others among the six genomes (Table S\ref{tab:genome}), we could not
obtain sufficiently long multiple sequence alignments with it, and had to
exclude it from the model-based analysis.  The estimated species tree using the
alignment of the five genomes (Figure S\ref{fig:mauve}) had the expected
topology \texttt{SC: The figure 1 must be nicer than one that now} (see Figure
\ref{fig:tree5}). \texttt{ClonalFrame} also estimated a relative strength of
recombination compared to mutations as $r/m=4.66$ meaning that four times more
sites had been affected by recombination-driven substitutions than by
mutation-driven substitutions (Table S\ref{tab:clonalframe}).  This indicated
potentially significant roles of recombination in the evolutionary process of
shaping these genomes.  Estimates of population mutation rate ($\theta$),
recombination rate ($\rho$), and average tract length ($\delta$) were obtained
for the alignment blocks analyzed (Figure S\ref{fig:scatter3}), and a median of
the estimates weighted over block lengths was obtained as a global estimate for
each of the three parameters (Table S\ref{tab:three}).  The local estimates
typically seemed to be moderately fluctuating around the global estimate, with
the exception of what appeared to be a recombination hotspot at a region 200
kilobase pairs from the origin of replication of SDE1.

The estimated species tree together with the global estimates of the three
population parameters were used to sample recombinant trees from the posterior
distribution in the second phase of \texttt{ClonalOrigin} \cite{Didelot2010}.
We summarized the posterior sample of recombinant trees by counting topologies
of the local genealogies induced by those recombinant trees for each of the
genomic positions (Table S\ref{tab:Gene-tree-topologies}).  Of the 105 possible
rooted topologies relatig five taxa, the dominant topology (66.7\%) was that of
the species tree.  The second most frequently sampled topology (9.3\%) was the
one in which SPY and SDE were sister clades to each other with SDD being as an
outgroup. Other topologies were sampled with frequency less than 4\%. This gave
initial indication to the extent of recombining gene transfer  between SPY and
SDE.

To gain a fine-scale view of the main types of recombination scenarios, we
looked at recombinant edges present in the posterior sample of recombinant
trees. Following the approach of Didelot et al. \cite{Didelot2010}, we
classified these edges by types defined by distinct pairs of donor and recipient
branches of the species tree (Figure \ref{fig:tree5}). For each edge type, we
averaged the number of edges present in the sampled recombinant tree, and
computed the ratio of this observed average frequency to a prior expected
frequency. The prior expected frequencies was computed by considering branch
lengths and their relative positions of the species tree, and the three
population parameter estimates in the preliminary step.  Base 2 logarithms of
these ratios are depicted in the heatmap of Figure
\ref{fig:Heatmap-of-recombination} (see also Tables S\ref{tab:heatmap} and
S\ref{tab:obsheatmap}).

Among the possible recombinant edge types, we can distinguish between ones that
preserve the species tree topology, and ones that alter it. We focused on 
topology-altering edge types both because our main objective was to study
cross-clade gene transfer, and because our simulation study indicated that the
observed counts of topology-preserving recombinant edges were unreliably
inferred (see \textbf{Simulations}). The heatmap allowed to find recombination
scenarios that deviated from the expectation dictated by the probabilistic
model.  We envisioned two main potential causes for such deviations: (1) reduced
or increased opportunities for species to interact and exchange genetic
material; (2) selection acting on newly transferred genes.

>>>>

% Potential reviewers and e-mail addresses 
% DANIEL FALUSH: d.falush@ucc.ie
% MATTHEW HAHN: mwh@indiana.edu 
% Debra Bessen: debra_bessen@nymc.edu
% Xavier Didelot: xavier.didelot@gmail.com
% Brian Golding: Golding@McMaster.CA

% TODO:
% Figure S\label{fig:calling-transfers} was not referred in the text.(Delete this?) 
% Figure S\label{fig:mowgli-sim} was not referred in the text. (Write comments)
% Table S\label{tab:sim-three} was not referred in the text. (Delete this?)
% 
%
% * Finalize section: Distribution of transfer events along the genome
% * Finalize section: Supp. Materials.
% Supp. Methods of Matt's modifications of Mowgli's method
% 
% Backup of the work
% swiftgen:/usr/projects/strep/gainloss/recombination

\documentclass[12pt]{article}

\usepackage{times}

@@

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}

@@

% Use doublespacing - comment out for single spacing
\usepackage{setspace}

% To rotate tables using sidewaystable instead of just table.
\usepackage{rotating}

@@

\usepackage{natbib}
\usepackage{../../latex/sty/genres}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
%\topmargin 0.0cm
%\oddsidemargin 0.5cm
%\evensidemargin 0.5cm
%\textwidth 16cm 
%\textheight 21cm

@@

@@

% FIXME: Remove the followings if you have to when submitting the manuscript.
\providecommand{\tabularnewline}{\\}
\newcommand{\lyxdot}{.}
\newcommand{\comment}[1]{}

% \@ifundefined{showcaptionsetup}{}{%
% \PassOptionsToPackage{caption=false}{subfig}}
% \usepackage{subfig}

@@

\pagestyle{myheadings}
%% ** EDIT HERE **
\markright{Horizontal gene transfer in Streptococcus}

@@

\begin{titlepage}

\title{Replacing and additive horizontal \\
gene transfer in {\em Streptococcus} }

\author{Sang Chul Choi$^{1}$, Matthew D.\ Rasmussen$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Ilan Gronau$^{1}$,
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

\date{ }
\maketitle

\begin{footnotesize}
\begin{center}
$^1$Department of Biological Statistics and Computational Biology,\\
Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^2$Department of Population Medicine and Diagnostic Sciences,\\
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\
\end{center}
\end{footnotesize}

\vspace{1in}

\begin{tabular}{lp{4.5in}}
{\bf Submission type:}& Research Article
\vspace{1ex}\\
{\bf Keywords:}&Bacterial evolutionary
genomics, recombination, {\em Streptococcus pyogenes}, {\em Streptococcus
  dysgalactiae} 
\vspace{1ex}\\
{\bf Running Head:}&Horizontal gene transfer in {\em Streptococcus}
\vspace{1ex}\\ 
{\bf Corresponding Author:}&
\begin{minipage}[t]{4in}
 Adam Siepel\\
 102E Weill Hall, Cornell University\\
 Ithaca, NY 14853\\
 Phone: +1-607-254-1157\\
 Fax: +1-607-255-4698\\
 Email: acs4@cornell.edu
\end{minipage}
\end{tabular}

\thispagestyle{empty}
\end{titlepage}

\doublespacing

% MBE limit is 350 words
\section*{Abstract}

The prominent role of horizontal gene transfer (HGT) in the evolution of
bacteria is now well documented, yet few studies have differentiated
between the mechanisms of homologous recombination and additive integration.
%as sources of HGT.  
These processes leave distinct phylogenetic signatures, because homologous
recombination causes a gene in one lineage to be replaced by a homolog from
another lineage (``replacing HGT''), while additive integration causes
addition without replacement (``additive HGT'').  Here we make use of these
signatures in a genome-wide investigation of HGT in the important human
pathogen {\em Streptococcus pyogenes} (SPY) and its close relatives {\em
  S.\ dysgalactiae} subspecies {\em equisimilis} (SDE) and {\em S.\
  dysgalactiae} subspecies {\em dysgalactiae} (SDD).  Using recently
developed statistical models and computational methods, we find evidence
for abundant gene flow of both kinds within the SPY and SDE clades, and of
generally reduced levels of exchange between SPY/SDE and SDD.  In addition,
our analysis strongly supports a previously reported, but unconfirmed,
finding of a pronounced asymmetry in SPY-SDE gene flow, favoring the
SPY-to-SDE direction, which may be associated with increasing virulence of
pathogenic SDE.  We find that this enrichment is driven by replacing
transfers, and likely reflects frequent homologous recombination between
co-occurring SPY and SDE cells in human hosts.  We find that virulence
genes are correlated with transfer events, but this correlation is driven
by additive, not replacement, HGTs.  The genes affected by additive HGTs
are enriched for transposition-related functions, while replacement HGTs
appear to influence a more diverse set of genes.  These findings shed new
light on the manner in which HGT has shaped pathogenic bacterial genomes
during relatively recent evolutionary time.

\thispagestyle{empty}
\clearpage

@@

During the last twenty years, it has become increasingly apparent that
horizontal gene transfer (HGT) has played a major role in genomic
evolution, particularly among bacteria and other microbes
\citep{Smith1992,Smith1993,Ochman2000,Koonin2001}.  Indeed, it is likely
that HGT has been sufficiently prevalent throughout evolutionary history
that no present-day gene can trace an unbroken history of vertical descent
to a common ancestor for all species \citep{Zhaxybayeva2011}.  This would
mean that no single phylogenetic marker is a reliable indicator of the tree
of life.

For HGT to occur in bacteria, DNA must first be transferred from a donor to
a recipient cell by transformation, transduction, or conjugation, then must
be integrated into the recipient cell's genome (setting aside the case of
replication-proficient plasmids).  This integration typically occurs in one
of two major ways: either the new sequence replaces a homologous sequence
through the process of homologous recombination (similar to ``gene
conversion'' in sexually reproducing organisms), or it is acquired through
an additive (non-replacing) integration process \citep{Thomas2005}.  These
two types of transfers leave distinct molecular evolutionary signatures and
will be referred to in this article as ``replacing'' and ``additive'' HGTs,
respectively (fig.\ \ref{fig:hgt}).  The rates at which these processes
occur depend on many factors, including the co-occurrence of bacterial
species in the environment,
%the propensity of donor cells to release intact DNA,
the degree of competence of recipient cells, sequence similarity between
the donor and recipient, and barriers to conjugation such as surface
exclusion.  In addition, the persistence of horizontally transferred DNA
segments in present-day species depends strongly on the subsequent effects
of natural selection.  Not surprisingly, rates of HGT appear to vary
considerably across groups of bacteria \citep[e.g.,][]{Feil2001}.

HGT in the {\em Streptococcus} genus is of particular interest.  This group
of Gram-positive bacteria contains the human pathogens {\em S.\ pyogenes}
(the cause of Group A streptococcal infections, including pharyngitis,
impetigo, cellulitis, necrotizing fasciitis, and rheumatic fever), {\em S.\
  pneumoniae} (bacterial pneumonia), {\em S.\ agalactiae} (neonatal sepsis),
{\em S.\ mutans} (dental caries), and {\em S.\ dysgalactiae} ssp.\ {\em
  equisimilis} (cellulitis, peritonitis, pneumonia, and other infections),
as well as the agricultural pathogens {\em S.\ dysgalactiae} ssp.\ {\em
  dysgalactiae} (mastitis in cows, ewes, and goats), {\em S.\ equi} ssp.\
{\em equi} (strangles in horses), and {\em S.\ canis} (various infections
in dogs and other animals).  Many of these species are concentrated in the
predominantly beta-hemolytic {\em Pyogenes} group.  Despite their
relatively recent evolutionary divergence, these {\em Pyogenes}-group
species have adapted to a remarkable variety of distinct ecological niches.
Accordingly, they display high rates of recombination and HGT, as well as
substantial evidence of positive selection
\citep{Feil2001,Marri2006,Anisimova2007,Lefebure2007,Suzuki2011}.  Complete
genome sequences are now available for several species in this group
\citep[e.g.,][]{Ferretti2001,Holden2009,Shimomura2011,Suzuki2011},
providing a rich resource for the study of the genomic evolution of
pathogenic bacteria over relatively recent time scales.

Within the {\em Pyogenes} group, several studies have focused on the issue
of HGT between the human-colonizing species {\em S.\ pyogenes} (here
denoted SPY) and {\em S.\ dysgalactiae} ssp.\ {\em equisimilis} (SDE).
Until recently, SDE was considered to be primarily a commensal organism
\citep{Vandamme1996}, unlike SPY, but it has become increasingly apparent
that it also has an important pathogenetic role, with a disease spectrum
similar to that of SPY \citep{Brandt2009}.  Shared virulence genes are well
documented between the two species \citep[e.g.,][]{Davies2007a}, genetic
exchange presumably having been enabled both by their close evolutionary
relationship and shared ecological niches, but
% Towers et al 2004
% Maxted and Potter, 1967
% Simpson et al., 1992
% Sriprakash et al., 1996
in a widely noted phylogenetic study of seven loci and more than 200
isolates, \cite{Kalia2001} also found extensive evidence of SPY-SDE gene
flow among housekeeping genes.  Interestingly, they observed a pronounced
bias in the direction of gene flow, favoring the SPY-to-SDE direction.
Gene flow has been confirmed in subsequent studies
\citep{Kalia2004,Davies2005,Davies2007a,Davies2007}, but the evidence for
directional asymmetry is mixed, with some indications of gene flow in the
opposite (SDE-to-SPY) direction \citep{Sachse2002}.  This issue has been
further complicated by the eventual retraction of Kalia et al.'s
\citeyearpar{Kalia2001} paper, due to the authors' inability to replicate
their findings and concerns about sample contamination.  The authors argued
that whole-genome sequencing may be required to definitively resolve open
questions about gene flow between these species.  These issues are of
particular interest because of the possibility of increasing virulence of
SDE due, at least in part, to gene flow from SPY.

Most recent studies of gene flow in bacteria have been based on relatively
simple phylogenetic methods that make use of incongruity of inferred gene
trees across loci \citep[e.g.,][]{Kalia2001,Kalia2004,Ahmad2009}.  These
methods are sensitive to errors in phylogenetic reconstruction, and ignore
information from branch lengths and from phylogenetic correlation at
adjacent loci.  Recently alternative, model-based approaches have been
proposed for the case of homologous recombination (``replacing HGTs'')
\citep{Didelot2007,Didelot2010}.  Using Bayesian principles and Markov
chain Monte Carlo (MCMC) techniques, these methods allow for uncertainty in
the phylogeny, make use of branch lengths and correlations across loci, and
allow for inference not only of individual recombination events but also of
parameters describing global rates and patterns of recombination.  In
addition, several recent methods have been introduced to detect HGT in a
phylogenetic framework, by parsimoniously reconciling reconstructed gene
trees with a given species tree, allowing for duplication, transfer, and
loss (DTL) events \citep{Merkle2010,David2011,Doyon2011,Doyon2011b}.  These
new model- and parsimony-based methods are complementary in many respects,
and could be particularly useful in combination.

In this article, we take a fresh look at the issue of HGT in the {\em
  Pyogenes} group, making use of newly available complete genome sequences
for SPY, SDE, and the related species {\em S.\ dysgalactiae} ssp.\ {\em
  dysgalactiae} (SDD), and new statistical and parsimony-based methods for
analysis.  The combined use of these methods allows us to examine both
replacing and additive HGTs, and compare their genome-wide effects.  We
find evidence for abundant gene flow within SPY, within SDE, and between
SPY and SDE, and for greatly reduced exchange between SPY/SDE and SDD.  In
addition, our genome-wide analysis supports Kalia et al.'s
\citeyearpar{Kalia2001} previously unconfirmed finding of a pronounced
preference for the SPY-to-SDE direction in SPY-SDE gene flow.
Interestingly, this property appears to hold for replacing but not for
additive transfers.  More generally, we find evidence for an association
between virulence genes and additive, but not replacing, HGTs.  Our results
have been made available as tracks in our recently released {\em
  Streptococcus} genome browser.

\section*{Methods}

\subsection*{Genome sequences and alignments}

Our primary analysis was based on five complete genome sequences, including
two representatives of {\em S.\ pyogenes} (accessions NC\_004070
[\citealp{Beres2002}; SPY1] and NC\_008024 [\citealp{Beres2006}, SPY2]),
two of {\em S.\ dysgalactiae} ssp.\ {\em equisimilis} (CP002215
[\citealp{Suzuki2011}; SDE1] and NC\_012891 [\citealp{Shimomura2011};
SDE2]), and one of {\em S.\ dysgalactiae} ssp.\ {\em dysgalactiae}
(AEGO00000000 [\citealp{Suzuki2011}; SDD]).  In addition, we used 
\textit{S.\ equi} ssp.\textit{\ equi} strain 4047 (NC\_012471
[\citealp{Holden2009}; SEE]) as an outgroup in our parsimony-based
analysis (supplementary table S\ref{tab:genome}).

For the model-based analysis of replacing transfers, we obtained
high-quality alignments of the five genomes (SPY1 \& 2, SDE1 \& 2, and
SDD) using the 
pipeline detailed by Didelot and colleagues
\citep{Didelot2007,Didelot2010}.  Briefly, we first aligned the genomes
using progressiveMauve v2.3.1 \citep{Darling2004,Darling2010} with default
options (supplementary fig.\ S\ref{fig:mauve}), and then used stripSubsetLCBs (from the
ClonalOrigin package; \citealp{Didelot2010}) to identify 276 blocks of
sequence alignments that exceed a length threshold of 1500 bp.  This
conservative threshold was selected to avoid biases from boundary effects
in short alignment blocks, based on a preliminary analysis.  Two blocks
containing long gaps were excluded from the analysis.
The final set of 274
alignment blocks was 1147632 nucleotide long.
Of 1951 genes of SPY1 genome 1125 genes were covered by the blocks.
Note that this alignment
procedure effectively focused on the ``core genome,'' by discarding regions
that were not conserved among the five species in question.  The 1125 genes
contained 0.17\% of gaps on average.

For the parsimony-based analysis, we began with 2314 gene family
clusters representing the six genomes (including the SEE outgroup)
from \cite{Suzuki2011}.  We aligned protein sequences corresponding to
each family using MUSCLE \citep{Edgar2004a}, and then obtained
nucleotide alignments by reverse translation, using the genomic
sequences as a guide.  Next we identified likely intragenic
recombinations using the Single Breakpoint Recombination (SBP) method
from the HyPhy package \citep{KosakovskyPond2006}.  Based on the
Akaike Information Criterion (AIC), SBP identified at least one
topology-altering recombination in 47.5\% (1094) of gene
families. These alignments were split into putatively non-recombining
gene fragments.  The final set of alignments contained 3408 gene
fragment families, 1794 of which contained one gene in each of the six
species.  Families ranged as small as two genes (417 fragments) to as
large as 59 (2 fragments) (supplementary fig.\ S\ref{fig:famsizes}).  Notably, in
contrast to the genome-wide alignments produced by progressiveMauve,
these alignments were representative of the ``pan genome,'' including
regions that frequently turn over, rather than just the ``core
genome'' \citep[e.g.,][]{Tettelin2005,Lefebure2007}.

\subsection*{Model-based analysis of replacing HGTs}

{\bf ClonalFrame and ClonalOrigin.}  To study replacing HGTs, we made use
of the coalescent-based model of homologous recombination and the Markov
chain Monte Carlo (MCMC) algorithm recently developed by
\cite{Didelot2010}.  The method was implemented in a 
program, called ClonalOrigin, that could produce samples from the posterior
distribution of {\em recombinant trees} given a set of alignment blocks. A
recombinant tree consists of a rooted tree with branch length, called a clonal
frame, for the five bacterial genomes,
which is estimated in a pre-processing step, augmented by a set of
{\em recombinant edges}. Each recombinant edge is defined by the two points
along the branches of clonal frame corresponding to the {\em donor} and
{\em recipient} in the hypothesized recombination event, and by the start
and end of the genomic interval that is affected by the event.  The donor
must pre-date the recipient, and a delayed coalescence is possible
(see \textbf{Supplementary Material} for details of summarizing recombinant
trees and edges). 
Another program, called ClonalFrame
\citep{Didelot2007}, can be used to infer the phylogeny that serves as the
basis of the recombinant trees.  ClonalFrame makes use of a similar, but
somewhat simpler probabilistic model, which permits inference
of the clonal frame. % and a set of model parameters.

{\bf Fitting the model.}  We used
ClonalFrame (v1.1) to estimate 
a clonal frame for the five-way
alignments, running the algorithm for $10^4$ burn-in iterations followed by
another $10^4$ sampling iterations.  A series of seven additional sampling
runs with 
random initializations converged to an essentially identical clonal frame, % model parameters,
indicating stable convergence.  We then used an initial run of
ClonalOrigin (subversion r19) to estimate global parameters including
the mutation rate, recombination rate, and recombinant tract length. We
applied the method separately to each of the 274 blocks, in all cases using
$10^6$ burn-in iterations, $10^7$ sampling iterations, and sub-sampling
every $10^5$ iterations. A global estimate for each parameter was then
obtained by taking the block-length weighted median of the block-specific
posterior means, which down-weighed less reliable estimates obtained from
shorter alignment blocks.  We generally followed \cite{Didelot2010} in 
fitting the model. % , with some minor differences.
